A Study on Common Bug Fix Patterns in Rust

### 2022-09-21 sections/methodology.tex
2.1.3

dictdiff provides modifications' types: 
\begin{itemize}
\item `add': A new structure has been added to the tree; 
\item `remove': A structure has been dropped from the tree; 
\item `change': The content of a sub structure of the first tree has changed.
\end{itemize}
dictdiff also determines the context within which the modification has occured. This context happens to be the root elements in our file, e.g. function blocks, struct blocks, impl blocks, etc. The modified content within a context is still in the form of tree. Our objective right now is to find primary elements in our diff that describe the semantics of occurred change.
% TODO rework the above paragraph

@@

The output of dictdiff is a list of diffs, where each diff comprises three parts. The first part specifies whether the modification is of type 'add', 'remove', 'change'. The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change (terminals and nonterminals). Our tool looks for changes within each root-level scope. That is, our assumption is that a change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool would detect both continuous and non-continuous line changes within one scope.
% TODO provide an example and then transform it into fig:extraction.

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one contributing element to the change. Through a Depth First Search we can collect all of these paths. We store only the nodes that exist in the \verb+DI+. The outcome of the DFS search is a set of paths.

Figure~\ref{fig:extraction} shows a simple change and its respective paths. The change replaces \verb+tail+ with \verb+&tail+ and \verb+head+ with \verb+head.clone()+. That is, after the change, the second argument passes a borrow of \verb+tail+ rather than sending the ownership of \verb+tail+ to the callee; and, in the third argument, instead of sending \verb+head+ to the callee, it sends a clone of it. A DFS search on this tree would yield three different paths, which we show in red, green, and purple. The path represents the sequence of involved non terminals, which represents one contributing factor to the change. We found it crucial to record the order of all path nodes as it gives us freedom to control the order-sensitiveness in our datapoint encoding.

The next question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. The number of columns would depend on the amount of information we want to encode in the datapoints. The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an encoding, for instance, there is no difference between two nested if statements and two if statements beside each other. Theoretically, to reach full order-sensitiveness, we would need encode all possible combinations of dimension items as our dimensions, which yields a large number (A loose upper bound would be $n^n$ where $n$ is the number of dimension items). Apart from that, not all combinations adhere to Rust syntax. A reasonable work around would be to reduce from the full set of nonterminals to a smaller set of categories (e.g. A category for larger entites like class or function definitions, and a category for smaller entities like statements and expressions), and then to set the dimension as all possible combinations of these categories (Similar to JS paper, ref). This would reduce the number of dimensions and provide more reasonable and syntactically-correct combinations. However, this would still yield a sparse dataset.

In our methodology, we propose a novel method. Similar to the first approach, we define the columns of our dataset, as the set of dimension items we collected from Syn. However, to achieve an acceptable level of order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of dimension items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

Our main objective is that our datapoints manifest the change with high level of precision. For instance, the purple (right-most) path in Figure~\ref{fig:extraction} could be described as a change in a local variable declaration; a change in a function call; calling a method of one of the function arguments; or calling clone() on one of the function arguments. The last description offers the most accurate description. The whole idea behind our weighting scheme is to be able to identify the change accurately.

% I need to reread this key paragraph. - PL
A simple heuristic is that, the closer we get to the leaves in the ASTDiff, the more important dimension items get. Another observation is that the elements closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). Exploiting the latter observation, to obtain a basic knowledge about the number of repetitions of dimension items, we mined the last 20 bug related commits of the last 20 most starred rust projects on GitHub, and ran them through of pipeline. We recorded the number of occurrences of each dimension item in total. Then we gave the value of the inverse of the number of occurrences to each token. The intuition is that the higher the number of occurrences of a dimension item, the closer it is usually seen to the root. Hence, it probably does not provide us with the most accurate description of the change. In addition to that, as we wanted to make sure that our tool captures the patterns related to borrow checker, we manually raised the weights of the dimension items related to borrow checker. That is why we characterize our weighting scheme as semi-automatically designed.

>>>>

An ASTDiff, as computed by dictdiff, contains a type:
\begin{itemize}
\item `add': A new structure has been added to the tree; 
\item `remove': A structure has been dropped from the tree; 
\item `change': The content of a sub structure of the first tree has changed.
\end{itemize}
% the description of the context in 2.1.2 is probably enough, you don't need it here, just remove it. I think that you just put the above itemize into 2.1.2 and erase the next paragraph, unless you can spot something you need to add to 2.1.2.
It also encodes the context of the modification. This context happens to be the root elements in our file, e.g. function blocks, struct blocks, impl blocks, etc. The modified content within a context is still in the form of tree. Our objective right now is to find primary elements in our diff that describe the semantics of occurred change.
% TODO rework the above paragraph

@@

% remove redundancy between the description of the ASTDiff in 2.1.1 and what we have here. Probably it can go all into 2.1.2, i.e. you can just move the last para of 2.1.1 to here and incorporate it.
The output of dictdiff is a list of diffs, where each diff comprises three parts. The first part specifies whether the modification is of type `add', `remove', or `change'. The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change (terminals and nonterminals). Our tool looks for changes within each root-level scope. That is, our assumption is that a change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool would detect both continuous and non-continuous line changes within one scope.
% TODO provide an example and then transform it into fig:extraction.

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one contributing element to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCL}$).

Figure~\ref{fig:extraction} illustrates path extraction on a simple change. The change replaces \verb+tail+ with \verb+&tail+ and \verb+head+ with \verb+head.clone()+. That is, after the change, the second argument passes a borrow of \verb+tail+ rather than sending the ownership of \verb+tail+ to the callee; and, in the third argument, instead of sending \verb+head+ to the callee, it sends a clone of it. A DFS on this tree yields three different paths, which we show in red, green, and purple. A path represents a sequence of involved non terminals. We found it crucial to record the order of nodes in the path, as this gives us the freedom to later record the order in our datapoint encoding.

% I think I would move the discussion of why fixed size from the beginning of the section to here; we don't need it at the beginning of the section.
% I think it would be better to show an example of the two if statements either way
\paragraph{Encoding the data.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. The number of columns would depend on the amount of information we want to encode in the datapoints. The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an encoding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to encode all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCL}$ as our dimensions, which yields an infeasible $O(n^n)$ dimensions; this could potentially be reduced somewhat, but is still impractical. A reasonable work around would be to reduce $n$ from the full set of non terminals to a smaller set of categories (e.g. a category for larger entites like class or function definitions, and a category for smaller entities like statements and expressions), and then to record all possible combinations of these categories (similar to JS authors~\cite{foo}). That approach reduces the number of dimensions and provides more reasonable and syntactically-correct combinations. However, it would still yield a sparse dataset. 
% todo you need to dig up the JS paper ref

% ok, add a figure and then we'll look at the text again and follow along with the picture
In our methodology, we propose a novel representation. Similar to the first approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCL}$ we collected from Syn. However, to simulate order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

Our main requirement is that our datapoints summarize the key changes in ASTDiffs---a datapoint should foreground the most important change in a diff. For instance, the purple (right-most) path in Figure~\ref{fig:extraction} could be described as a change in a local variable declaration; a change in a function call; calling a method of one of the function arguments; or calling clone() on one of the function arguments. The last description is the most useful one, and we designed our weighting scheme to encode this value judgment.

We made two observations about changes. First, we saw that an item (non terminal or BC-related lexeme) that occurs closer to AST leaves tends to be more semantically important. Relatedly, items closer to the root tend to be repeated in all the paths within one ASTDiff (e.g. ItemFn, Local, ExprCall in Figure~\ref{fig:extraction}). We thus want to prioritize items that tend to occur closer to leaves and de-prioritize items that tend to occur closer to roots.

To design our weighting scheme, we collected empirical data about locations of item occurrences in a corpus. Specifically, we mined the last 20 bug related commits of the last 20 most-starred Rust projects on GitHub, and ran them through our pipeline. We recorded the total number of occurrences of each item. Then, if $\# i$ is the number of occurences of $i$, we gave $i$ a weight of $1/\# i$. Because items that occur closer to the root also occur more often, per our observation, this gives higher weights to items that occur infrequently, i.e. closer to the leaves. Items that occur often and have lower weights are, we believe, less important in describing a change.

Furthermore, as we wanted to make sure that our tool captures patterns related to the Rust borrow checker, we manually increased the weights of the \texttt{BCL} items. The manual adjustment is why we characterize our weighting scheme as semi-automatically designed.
